Why do languages tolerate heterography? An experimental investigation of the cultural evolution of informative orthography

### 2023-08-29 manuscript/wp4.tex
ms 25/08

The results from all ten chains (labelled A--J) in the Transmission-only condition are shown in Fig.~\ref{dif_lrn}. Each 3×3 matrix represents the suffix spelling system in use at a particular generation with shape down the rows and color along the columns, just as in Fig.~\ref{stimuli}. For example, the system at Generation~9 in Chain~D uses three spellings---<koe>, <ko>, and <co>---to express the shape dimension. We describe such a system as ``redundant'' because shape is consistently expressed by the stem, so the suffix spellings that emerged in this case convey no additional information beyond what is already conveyed by the stem. Similar outcomes occurred in Chains~C, E, F, and~H and are characteristic of the Transmission-only condition. We also saw the emergence of fully transparent suffix spelling systems in Chains~A, B, I, and~J. Chain~I, for example, ultimately uses a single spelling, <coe>, to represent the /kəʊ/ sound; that is, the written suffix forms make no distinction between shapes or colors, just like the spoken language. Chain~G did not settle on a clear pattern, using <co>, <coe>, and <coh> somewhat interchangeably in the final generation, although there were some signs of a color-expressive system emerging in, for example, Generation~6, where yellow is consistently spelled <coh>, blue is consistently spelled <coe>, and pink uses both <co> and <coe>. The only other signs of color-expressive systems are H1, which was rapidly converted into a redundant system in subsequent generations, and J2, which ultimately degenerated towards transparency.

@@

As a quantitative measure of how informative the orthographic systems are over time, we adopt communicative cost \parencite{KempRegier:2012, Regier:2015, Kemp:2018}. Communicative cost expresses how much information is lost, on average, when a speaker/writer attempts to convey a meaning to a listener/reader using some shared signaling system. If the system contains no ambiguities (all meanings are expressed by unique signals), communicative cost will be zero---that is, zero information will be lost during each attempt to communicate using that system. Communicative cost will take some larger value if the system contains ambiguity. Formally, it is given by $\sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) )$,
	% \begin{equation}
	% \sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) ),
	% \end{equation}
where $U$ is the universe of meanings that may be expressed, $\mathrm{Pr}(m)$ is the probability that a particular meaning would need to be expressed, and $\mathrm{Pr}(m|s_m)$ is the probability that a reader would infer meaning $m$ given that a writer produced signal $s$ for meaning $m$. In our case, $U$ is the set of nine alien objects, $\mathrm{Pr}(m)$ is set to $1/|U|$ (all objects need to be talked about with equal probability), and $\mathrm{Pr}(m|s_m)$ is given by $1/|M_s|$, where $M_s$ is the set of meanings labeled $s$ according to the system.

@@

To analyze more formally what types of system tend to emerge under different conditions, we first designated four reference systems that are of particular interest: holistic,\footnote{Although we use the term holistic here, this category also encompasses compositional suffixes that express both shape and color in a systematic way (specifically, K9, M1, M7, and Q7).} expressive, redundant, and degenerate, as illustrated in Fig.~\ref{}. These four typological categories form a continuum in terms of the number of suffix forms they make use of and how these forms are conditioned on meaning. The holistic system has nine unique forms that express both color and shape, the expressive and redundant systems have three unique forms that express either color (expressive) or shape (redundant), and the degenerate category has one unique form that expresses nothing. To classify a participant's output into one of these four categories, we may compute which of the reference systems is most similar in terms of its information content. To do this, we formalize the systems as set partitions (i.e., each system may be viewed as a partitioning of the universe of meanings into $n$ disjoint sets) and adopt variation of information \parencite[VoI;][]{Meila:2007} as a distance metric between any two such partitions.\footnote{VoI is a proper metric on set partitions, measuring the amount of information (in bits) that is lost and gained in the transformation of one set partition into another.}

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/typ_dist_dif.eps}}
	\vspace*{2pt}
	\caption{Typological distribution at each generation and in each condition across four typological categories: holistic (H; purple), expressive (E; blue), redundant (R, green), and degenerate (D; yellow). At generation~0, all 10 chains in both conditions use a holistic system (by design). In Transmission-only, the systems switch first to a redundant and then to a degenerate patterning. In Transmission + Communication, the systems ultimately become degenerate, but also tend to favor holistic systems over redundant ones along the way. Expressive systems were rare in both conditions.}
	\label{typ_dist_dif}
	\end{figure}

Rather than simply classify each system into exactly one typological category, we computed probability distributions over the typological space according to distance. This gives a slightly more nuanced view of the typological distribution by accounting for cases where a given system might be partially redolent of one category but also redolent of another (for example, system T9 is part redundant and part degenerate). The probability distribution was calculated as $e^{-3\cdot\mathrm{VOI}(p, q)}$

Fig.~\ref{typ_dist_dif} plots the typological distribution (i.e., across chains) for each generation and each condition.

We computed the VoI between the emergent suffix systems and the three reference systems and plotted them on a ternary diagram, the three vertices of which represent the three reference systems (see Fig.~\ref{exp1_ternary_plot}). The diagram is divided into three sectors (shades of gray), which reveal which reference system the final converged-on suffix systems are most similar to. Most of the emergent system are either redundant or degenerate, with some straddling the border between the two. No system is classified as expressive, although two of the converged-on systems (highlighted in Fig.~\ref{exp1_ternary_plot}) showed some signs of color-marking. In general, there is no discernible difference between the two conditions; regardless of whether the communicative pressure was present, the emergent suffix spelling systems tended to fall along the degenerate--redundant axis, and the only two cases of color-marking were in~fact from the Transmission-only condition.

>>>>

The results from all ten chains (labelled A--J) in the Transmission-only condition are shown in Fig.~\ref{dif_lrn}. Each 3×3 matrix represents the suffix spelling system in use at a particular generation with shape down the rows and color along the columns, just as in Fig.~\ref{stimuli}. For example, the system at Generation~9 in Chain~D uses three spellings---<koe>, <ko>, and <co>---to express the shape dimension. We describe such a system as ``redundant'' because shape is always expressed by the stem, so the suffix spellings that emerged in this case therefore convey no additional information. Similar outcomes occurred in Chains~C, E, F, and~H and are characteristic of the Transmission-only condition. We also saw the emergence of fully transparent suffix spelling systems in Chains~A, B, I, and~J. Chain~I, for example, ultimately uses a single spelling, <coe>, to represent the /kəʊ/ sound; that is, the written suffix forms make no distinction between shapes or colors, just like the spoken language. Chain~G did not settle on a clear pattern, using <co>, <coe>, and <coh> somewhat interchangeably in the final generation, although there were some signs of a color-expressive system emerging in, for example, Generation~6, where yellow is consistently spelled <coh>, blue is consistently spelled <coe>, and pink uses both <co> and <coe>. The only other signs of color-expressive systems are H1, which was rapidly converted into a redundant system in subsequent generations, and J2, which ultimately degenerated towards transparency.

@@

To analyze more formally what types of system tend to emerge under different conditions, we first designated four reference systems that are of particular interest: holistic,\footnote{Although we use the term holistic here, this category also encompasses compositional suffixes that express both shape and color in a systematic way (specifically, K9, M1, M7, and Q7).} expressive, redundant, and degenerate. These four typological categories may be positioned along a complexity/cost continuum in terms of the number of suffix forms they make use of and how these forms are conditioned on meaning, as illustrated in Fig.~\ref{}. The holistic system has nine unique forms that express both color and shape, the expressive and redundant systems have three unique forms that express either color (expressive) or shape (redundant), and the degenerate category has one unique form that expresses nothing. To classify a participant's output into one of these four categories, we compute which of the reference systems is most similar in terms of its information content. To do this, we formalize the systems as set partitions (i.e., a partitioning of the universe of meanings into disjoint subsets) with variation of information \parencite[VoI;][]{Meila:2007} defined as the distance metric between any two such partitions.\footnote{VoI is a proper metric on set partitions, measuring the amount of information (in bits) that is lost and gained in the transformation of one set partition into another. Under this metric, the holistic and degenerate systems are considered very dissimilar because they carry very different levels of information content. The expressive and redundant systems are also considered very dissimilar because, although they carry the same amount of information, the information they carry is orthogonal (shape in the case of redundant, and color in the case of expressive).} Then, rather than simply classify a given system into exactly one typological category, we compute a probability distribution over the typological space according to the respective distances of the four reference systems. This gives a slightly more nuanced view of the typological distribution by accounting for cases where a given system might be partially suggestive of one typological category but also suggestive of another (e.g., T9 is part redundant, part degenerate). The typological probability distribution $t$ over the set of four reference systems $R$ was calculated for participant system $s$ according to% $t(i) \propto e^{-2\mathrm{VoI}(R_i, s)^2}$,
\begin{equation}
	t(i) \propto e^{-2 \cdot \mathrm{VoI}(R_i, s)^2}
\end{equation}
and averaged over chains for a given generation.

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/typ_dist_dif.eps}}
	\vspace*{2pt}
	\caption{Typological distribution at each generation and in each condition across four typological categories: holistic (H; purple), expressive (E; blue), redundant (R, green), and degenerate (D; yellow). At Generation~0, all 10 chains in both conditions use a holistic system (by design). In Transmission-only, the systems switch first to a redundant and then to a degenerate patterning. In Transmission + Communication, the systems ultimately become degenerate, but also tend to favor holistic systems over redundant ones along the way. Expressive systems were rare in both conditions.}
	\label{typ_dist_dif}
	\end{figure}

The results are plotted in Fig.~\ref{typ_dist_dif}. These probability distributions can be read as the proportion of chains that adopted each of the four configurations at each generation. At Generation~0, all chains are holistic by design. In Transmission-only, there is an initial rapid increase in the number of chains classified into the redundant typological category. The dominance of the redundant category is then gradually eroded as the chains transition to degeneracy. In Transmission + Communication, there is initially a fairly even mix of all four kinds of system, but by Generation~9, degenerate systems tend to be most common. There is, however, a notable increase in holistic systems emerging in later generations; these are the cases of compositional suffixes that arose through the addition of English color letters on top of redundant suffix spellings.

As a quantitative measure of how informative the orthographic systems are over time, we adopt communicative cost \parencite{KempRegier:2012, Regier:2015, Kemp:2018}. Communicative cost expresses how much information is lost, on average, when a speaker/writer attempts to convey a meaning to a listener/reader using some shared signaling system. If the system contains no ambiguities (all meanings are expressed by unique signals), communicative cost will be zero---that is, zero information will be lost during each attempt to communicate using that system. Communicative cost will take some larger value if the system contains ambiguity. Formally, it is given by% $\sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) )$,
	\begin{equation}
	\sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) ),
	\end{equation}
where $U$ is the universe of meanings that may be expressed, $\mathrm{Pr}(m)$ is the probability that a particular meaning would need to be expressed, and $\mathrm{Pr}(m|s_m)$ is the probability that a reader would infer meaning $m$ given that a writer produced signal $s$ for meaning $m$. In our case, $U$ is the set of nine alien objects, $\mathrm{Pr}(m)$ is set to $1/|U|$ (all objects need to be talked about with equal probability), and $\mathrm{Pr}(m|s_m)$ is given by $1/|M_s|$, where $M_s$ is the set of meanings labeled $s$ according to the system.

@@
